Case-based Arabic Morphological Analyzer for Arabic Text Mining

### 2010-12-17 trunk/tex/sarf.aclhlt.tex
final before 8

%\title{Instructions for NAACL HLT 2010 Proceedings\Thanks{This...}}
\title{Case-based Arabic Morphological Analysis }

@@

In this paper, we present Sarf, a {\em novel case-based efficient
morphological analyzer} that uses compositional 
non-deterministic Automata driven by a case based controller.
Each alive machine in Sarf represents a valid morphological analysis. 
Sarf keeps alive all possible analyses of the text and gives 
opportunity to the controller to intervene at every single input 
letter. 
The decisions of the case-based controller can thus prune false 
positives early in the run. 
Each alive machine in Sarf takes as input one letter at a time 
from a text stream. 
Sarf does not assume that the word at hand is a token and
performs tokenization on the fly based on morphological correctness.
Up to our knowledge, this is the first morphological analyzer that 
handles the ``run-on'' words problem.

Sarf uses a {\em recursive} concatenative analysis with a novel 
refinement. 
Sarf introduces recursive affixes in order to
retain better part of speech information and enhance the 
efficiency of affix matching.

\transfalse
\begin{figure}[tb]
\center{
\resizebox{1.2\columnwidth}{!}
{ \input{figs/exhadith.pdftex_t}}
\caption{Hadith abstraction example.}
\label{f:exhadith}
}
\end{figure}
% the FSMs for the three words
\transtrue

We validate our hypothesis using a case study from the Islamic 
literature. 
A \RL{.hady_t} is a narration related to the prophet Mohammad
through a \RL{sanad} or a sequence of narrators. 
Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in Arabic with its 
transliteration and translation. 
The authentication of a \noArRL{.hady_t} highly depends on the credibility
of each of the narrators as reported in separate biography 
books. 
In this paper we consider the problem of automatically segmenting
a \noArRL{.hady_t} book into narrations, then segmenting each narration into
its content or \RL{matn} and its \noArRL{sanad}.
We also partition the sanad accurately into the 
separate narrators so that we can later look each one of them 
up in the biography books. 
For example, the boxes in Figure~\ref{f:exhadith} are proper names and once
connected toghether they form complex names of narrators. For example
narrator $n_1$ has the first name \noTrRL{qtybT} and his father name is
\noTrRL{s`yd} as the word in between \noTrRL{bn} (son of) indicates.

The hadith case study is interesting to Sarf since the controller
considers most of the analyses where we have a concentration
of proper names in the text. It ignores the text where
proper names occur rarely as long 
as the text is valid.
The controller is also interested in discovering a subset of words 
that relate persons to each other such as \RL{bin} or ``the son of''
or words that mean narrate such as \RL{qaal} or ``said''. 
With Sarf, we successfully automated the analysis of 
three books of Islamic narrations~\cite{IbnHanbal,AlTousi,AlKulayni}
and extracted from each one of them a vector of a three level deep
complex structure. 
We also showcase that the results of the accurate high level abstraction
results of the case-based analysis was not affected by a major
refinement to the analyzer that increased low level analysis accuracy.

@@

Sarf is similar to Beesley~\shortcite{Beesley:01} in that it uses
finite-state Automata (FSA). 
Beesley reports that the number of the Automata generated by a compiler
from Xerox rules can not be controlled and also reports a difficulty to 
compose the FSAs into a single FSA framework. 
Sarf solves the problem with an FSA that is constructed from 
the controller and 
the prefix, stem, and suffix deterministic tansducers. 
Sarf differs in that it uses recursive affixes in the 
affix transducers, and in that the transducers consider
input from the controller. 
With Sarf it is easy to control the FSA with a custom user case 
controller. Up to our understanding, Beesley  
requires recompiling the Xerox rules.

Sarf is similar to The MADA+TOKAN toolkit in that it thinks of
the several valid morphological analyses as a richness that 
should be exploited at a higher abstract level rather than
an inaccuracy that should be corrected. 
Sarf differs in that it provides an interface for the 
case study to interfere in judging the analyses much earlier and
at a finer granularity level. 
Also in Sarf there is no need for a separate tokenizer such as
TOKAN as each alive Automata carries with it the positions in 
text where a composition decision was made. 
The case-based controller can build the trace 
when needed.

@@

We strongly feel that the refinement of SAMA~\cite{Maamouri:10} 
is the closest to our approach. SAMA was refined to interact with
the ATB~\cite{Maamouri:04} project after the addition of a large 
new corpus. We find supporting evidence in this decision to our hypothesis
about the utility of a case-based analyzer. 
The refinements that resulted in algorithmic changes in SAMA were 
done manually and worked in integration with the ATB format. 
Sarf is a case-based controller approach that can modify 
the behavior of the analyzer on the fly and can interact
with the any case study. In Sarf, the analyzer remains general, and the 
case-based controller specializes the analysis.

@@

Sarf encodes the prefix, suffix and stem lexicons into 
three linear transducers and uses a non-deterministic composition
of the three to compute all valid morphological analyses.
We first present a running example of Sarf. 
The running example explains
how the non-deterministic composition works and 
illustrates how Sarf implements recursive affixes and
handles ``run-on'' words.
%Then we discuss how Sarf builds and composes the different FSAs.

@@

\transfalse
\begin{figure*}[tb]
\center{
\resizebox{1.7\columnwidth}{!}
{ \input{figs/FSM.pdftex_t}}
\caption{Example affix and stem transducers.}
\label{f:example}
}
\end{figure*}
% the FSMs for the three words
\transtrue

The diagram in Figure~\ref{f:example}
illustrates the operation of Sarf when parsing the
string \RL{al-lA`bwn}\RL{wsyl`b-h-aa}.
\footnote{ 
\RL{n}%
\RL{w}
\RL{b}
\RL{`}
\RL{a} 
\RL{l}
\RL{l}
\RL{a} %
\RL{a}
\RL{h}
\RL{b}
\RL{`}
\RL{l}
\RL{y}
\RL{s}
\RL{w}
in separate form to ease following the example
}
The square and circle legends are states 
and the directed edges are transitions driven by an input
letter that matches the label of the edge. 
The square denotes an accept state, 
the circle denotes a regular state, and the reject states
are omitted for clarity. 
Figure~\ref{f:example}(a) represents the prefix transducer and 
Figures~\ref{f:example}(b) and~\ref{f:example}(c) represent the 
stem and suffix transducers respectively. 
The symbol $\epsilon$ represents an empty string and is 
the source of non-determinism in this example.

We start at the root square state in the prefix transducer
which is an accept state. 
The edge labeled $\epsilon$ connects the prefix 
transducer to the stem transducer and transitions from any accept state
in the prefix transducer to the root state in the stem transducer.
The edge that connects the stem transducer to the suffix transducer
follows the same behavior.

When there are two valid transitions such as \RL{w} 
and $\epsilon$ in the start case, the non-deterministic 
composition $\Phi$ spawns an exact copy of itself $\Psi$. 
$\Phi$ makes one transition, and $\Psi$ makes the other. 
Each of $\Phi$ and $\Psi$ now represent a valid analysis so far. 
When $\Phi$ or $\Psi$ is at a state where the next input letter
leads to a reject state, it dies.
The reject states are omitted from Figure~\ref{f:example}
and are denoted by the absence of the corresponding 
letters. 
In our example, if we assume there were no stems that start 
with the letter \RL{w}, $\Psi$ will die then. 
In reality probably $\Psi$ will die when the input string
is at \RL{wsyl`}.

Note that the prefix and affix transducers allow recursive 
prefixes. For example, \RL{w} will result in an accept state
that transitions to the stem transducer. 
When \RL{s} follows, we move to another accept state in the 
prefix transducer corresponding to the suffix \RL{ws}. 
The same applies to the suffix transducer.

Lets consider $\Phi$ after it consumed \RL{wsy} 
and transitioned into the stem transducer root node. 
Now $\Phi$ will transition with \RL{l`b} to reach an accept 
state. 
Before moving with the letter \RL{b} to the accept state,
$\Phi$ needs to make sure that the stem \RL{l`b} is compatible
with the prefix \RL{wsy}. 
It does that by associating a category value for each accept state. 
Sarf considers the value of the category as part of the state.
Thus each state shown in Figure~\ref{f:example} actually represents
more than one state. 
If the category of \RL{l`b} is compatible with the category of
\RL{wsy} then \RL{b} will move $\Phi$ to an accept state. 
Otherwise, it will move it to a regular or a reject state.

Since $\Phi$ is now in an accept state in the stem transducer, it will
spawn $\Sigma$ that transitions into the root of the 
suffix transducer. When \RL{h} is consumed, $\Phi$ will die since 
there is no edge from the current state labeled with \RL{h}. 
$\Xi$ represents a valid analysis and can proceed.

Note that $\Sigma$ reports a full word at any accept state
by spawning a new composition using the $\epsilon$ transition
to the prefix transducer.
In Sarf, this only happens with white space, delimiters, 
and non-connecting letters. 
We left that out in Figure~\ref{f:example} for clarity.

@@

\begin{figure*}[tb]
\center{
\resizebox{1.8\columnwidth}{!}
{ \input{figs/abstract_machine.pdftex_t}}
\caption{An abstraction of the non-deterministic construction of  
    Sarf from the affix and stem transducers.}
\label{f:composition}
}
\end{figure*}

@@

\begin{table}[tb]
\centering
\begin{tabular} {p{6cm}}
\begin{Verbatim}[fontsize=\relsize{-1},
frame=topline,framesep=4mm,label=\fbox{NDSarf algorithm},
commandchars=\\\{\}, codes={\catcode`$=3\catcode`_=8}]
NDSarf(string $L$, Controller $A$) 
  MachineVec $M$; -- collection of machines
  Machine $\Phi_1$;
  $M$.insert($\Phi_1$);
  foreach $c$ in $L$ \{
    foreach $\Phi$ in $M$ \{
      if ( $\Phi$.state.isAccept() ) \{
        if ($\Phi$.isSuffix())
          if ($c$.isWhite()) or 
            $A$.report();
          if ($\Phi$.lastChar().isNonConnecting())
            $A$.report();
        $\Psi$ = $\Phi$.clone();
        $M$.insert($\Psi$); \}

      if ( $\Phi$.isWalkable($c$) ) \{
        $\Phi$.transition($c$);
      \} else \{
        $M$.remove($\Phi$);
        $\Phi$.die();
      \} \} 
    $A$.control($M$, $c$); \}
\end{Verbatim}
\end{tabular}
\label{a:ndsarf}
\end{table}

@@

@@

\begin{figure}[tb!]
\center{
\resizebox{.9\columnwidth}{!}
{ \input{figs/hadith.pdftex_t}}
\caption{The controller of the hadith case study.}
\label{f:hadith}
}
\end{figure}

@@

\begin{table*}[bt]
\centering
\caption{Results of the hadith case study with Sarf.}
%\begin{tabular}{|p{1.5cm}||c|c||c|c||c|c|} \hline
\begin{tabular}{lp{.2cm}ccp{.2cm}ccp{.2cm}cc} %\cline{2-10}
 & & \multicolumn{2}{c}{Al Kafi} & & \multicolumn{2}{c}{Al Istibsar} & &\multicolumn{2}{c}{Ibn Hanbal} \\ \cline{3-10}
 & & Simple & Refined & & Simple & Refined & & Simple  & Refined \\\bottomrule  %\hline \hline
Word count & &\multicolumn{2}{c}{98,943} & & \multicolumn{2}{c}{103,835} & & \multicolumn{2}{c}{20,354} \\ 
 Names      & & 20,473 & 12,060  & &20,646 &  14,613& & 4,762 & 3,013\\
Names/Narrator& &2.48 & 1.97 & & 2.17 &  1.84& & 1.49 & 1.25 \\
Narrators & &3,080 & 2,623 & & 6,164 &  5,767& & 2,082 & 1,755 \\ 
Narrators/Chain & & 5.28 & 4.84 & & 4.98&  4.76 & & 4.8 &4.05 \\
Chains & & 583 & 542 &  & 1,238& 1,211& & 433 & 433 \\ 
Ignored names & & 11,801 & 6,400 &  & 5,845 & 3,348 & & 1,386 & 642 \\ \hline
Segmentation accuracy & & 85\% & 96\%& & 96\%& 96\%& & 92\%& 92\%\\ 
Chain accuracy & & 99\% & 99\%&  & 98\%& 99\%& & 99\% & 97\% \\ 
%Narrator accuracy & & & & & &  & & & \\ 
Narrator name accuracy & & 93\% & 91\%& &  92\%& 90\% & & 90\% & 90\% \\ \hline
Name false positives & & 28\% & 7\%&  & 9\%& 4\% & & 34\% & 4\% \\ \hline
Sarf running time (secs.)& & 2.97 & 1.32 & & 2.93 & 1.31 & & .25 & .096\\ \hline \hline
 & & Buckwalter& ElixirFM & & Buckwalter & ElixirFM & & Buckwalter & ElixirFM \\\bottomrule
Names      & & 18,354 & 19,201 & & 17,934 &  18,202 & & 5,021 & 5,117 \\
Names/Narrator& & 2.6 & 2.1 & & 2.47 &  2.07 & & 1.83 & 1.62\\
Narrators & & 3,210 & 2,981 & & 6,425 &  6,208 & & 2,311 & 1,965 \\ 
Narrators/Chain & & 5.57 & 5.32 & & 5.46 &  5.32 & & 5.12 & 5.01 \\
Chains & & 627 & 601 &  & 1,318 & 1,250 & & 487 & 452 \\ 
Ignored names & & 9,642 & 9,915 &  & 6,118 & 6,874 & & 1,812 & 1,904 \\ \hline
Segmentation accuracy & & 76\%& 84\%& & 85\%& 89 \%& & 84\%& 85\%\\ 
Chain accuracy & & 78\% & 87\%&  & 87\%& 90\%& & 85\% & 86\% \\ 
%Narrator accuracy & & & & & &  & & & \\ 
Narrator name accuracy & & 76\% & 86\%& & 84\%& 87\% & & 84\% & 84\% \\ \hline
Name false positives & & 29\% & 22\%&  & 14\%& 13\% & & 36\% & 32\% \\ \hline
Running time (secs.)& & 6.65 & 2.78$\times60^2$& &4.55 & 2.3$\times60^2$ & & .66 & 29.2$\times$60 
\end{tabular}
\normalsize
\label{t:hadithresallresults}
\end{table*}

In section we present results that compare Sarf to existing 
morphological analyzers such as Buckwalter and ElixirFM. 
We also evaluate our hypothesis of the utility of a case-based 
morphological analyzer,
by running our case study in two modes, simple and refined.
We ran our experiments using a dual core 2.66 Ghz 64-bit processor 
with 4GB of memory using a Linux operating system.

In the simple mode, Sarf is controlled by the hadith controller.
In the refined mode, we apply several refinements to Sarf.
The first refinement ignore female proper names except at the end of
a chain since they do not occur often in narrator names in hadith while
they can be common in early narrators such as the wives of the prophet.
We also add a refinement that detects compound proper names 
such as \RL{`bd alr.hmn} 
as they are common in narrator names.
Finally, we consider a possessive form as a reference to a person and treat
it as a narrator.

We report our results in Table~\ref{t:hadithresallresults} on the three hadith books.
The word count row is an indicator of the size of data.
The number of names, narrators, and chains reflect a hierarchical structure where a chain contains
narrators, and a narrator contains names.
In general, we observe that the refinements slightly improved performance and accuracy at
the lowest level of abstraction by reducing false positives.
However, the refinements had little to no effect on the high accuracy measures at higher
levels of abstraction.
We believe that this is a strong supporting evidence to our hypothesis.

The names row reports the total number of detected names. The names/narrator row reports the average
number of names per narrator structure. The narrators row reports the number of narrators detected.
The narrators/chain row reports the number of narrators per chain structure. And the chain row reports
the number of detected chains in the book.
The ignored names row reports the number of names detected outside the chains and thus irrelevant to
the case study.
The segmentation accuracy reports on the accuracy of segmentation of the book into separate hadiths.
The metric is computed manually over a representative sample of each text. The same applies to all other accuracy measures.
The chain accuracy reflects the percentage of correct narrators covered by extracted chains.
The narrator name accuracy row reflects the percentage of correct proper names detected within
narrator structures.

The name false positives row reports the percentage of words recognized as names while they should not.
We notice that the low level metric of name false positives improved dramatically with the refined
analysis without substantially affecting the high abstract measures such as chain and segmentation
accuracy.
We notice that the controller in the simple analysis ignored more names and thus the case-based
controller compensated for the inaccuracy of the low level analysis with a simple heuristic and
without the need of the fancy refinements.

We investigated the case of Al Kafi where the accuracy of the simple analysis was significantly lower than that of the refined
analysis. We  discovered that the chains detected by the refined analysis were a subset of those detected by the simple
analysis. 
The reason for the lower accuracy is false positive chains. 
The simple analysis did not miss important information. 
We hypothesize that a higher level of abstraction such as intersecting the narrators from the hadith books with the narrators 
in the biographies can reduce the false positives in the chains.

@@

We implemented our case study with Buckwalter in two phases. 
In phase one, we process the entire text with Buckwalter and
obtain annotations to locations in text. 
Then in the second phase we feed these annotations to 
our controller as if they were coming from Sarf. 
We measure the time it took to produce the annotations
and the time it took the controller to process the 
annotations whithout the time spent streaming the annotations.
We used the same setup for ElixirFM.

Sarf outperforms both Buckwalter and ElixirFM in runing time. 
It also produces higher accuracy measures. 
We think this is mainly due to the extended
lexicon as well as to the fact that Sarf reduces false positive
names by ignoring prefixes and suffixes that can not form
names even if the stem was a name. 
Doing the same with Buckwalter and ElixirFM requires  
expanding the annotations resulting from one affix to reflect
its multiple composing affixes.
This is easily done with the way we defined affixes recursively.

Buckwalter performed much better than ElixirFM. This is up to 
our knowledge the first comparison of these two tools and 
we think the big difference is due to the more exact 
and grammar based functional morphological analysis ElixirFM 
uses.

>>>>

%\title{Instructions for NAACL HLT 2010 Proceedings\Thanks{This...}}
\title{Case-based Arabic Morphological Analyzer for Arabic Text Mining}

@@

In this paper, we present Sarf\footnote{Sarf (name modified for blind review purposes) 
will be available online as an open source tool.},
 a {\em novel case-based efficient
morphological analyzer} that uses 
non-deterministic finite state Automata 
driven by a case based controller.
%Each alive machine in Sarf represents a valid morphological analysis.
Sarf keeps alive all possible analyses of the text and gives 
opportunity to the controller to intervene at every single input 
letter. 
The decisions of the case-based controller can thus prune false 
positives early in the run. 
Each alive machine in Sarf takes as input one letter at a time 
from a text stream. 
Sarf does not assume that the word at hand is a token and
performs tokenization on the fly based on morphological correctness.
Up to our knowledge, this is the first morphological analyzer that 
handles the ``run-on'' words problem.

Sarf uses concatenative analysis with a novel 
{\em recursive affixes} refinement. 
Sarf introduces recursive affixes in order to
better retain part of speech information and enhance the 
efficiency of affix matching.

\transfalse
\begin{figure}[tb]
\center{
\resizebox{.9\columnwidth}{!}
{ \input{figs/exhadith.pdftex_t}}
\caption{Hadith abstraction example.}
\label{f:exhadith}
}
\end{figure}
% the FSMs for the three words
\transtrue

We validate our hypothesis using a case study from the Islamic 
literature. 
A \RL{.hady_t} is a narration related to the prophet Mohammad
through a \RL{sanad} or a sequence of narrators. 
Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in 
Arabic with its transliteration and translation. 
The authentication of a \noArRL{.hady_t} depends on 
the credibility of each of the narrators as reported in 
separate biography books. 
In this paper we consider the problem of automatically segmenting
a \noArRL{.hady_t} book into narrations, then segmenting each 
narration into its content or \RL{matn} and its \noArRL{sanad}.
We also partition the sanad into the 
separate narrators so that we can later look each one of them 
up in the biography books. 
For example, the boxes in 
Figure~\ref{f:exhadith} are proper names and once
connected together they form complex names of narrators. 
For example, narrator $n_1$ has the first name 
\noTrRL{qtybT} and his father name is
\noTrRL{s`yd} as the word in between \noTrRL{bn} (son of) indicates.

The hadith case study is interesting to Sarf since the controller
considers most of the analyses where we have a concentration
of proper names in the text. It ignores the text where
proper names occur rarely.% as long 
%as the text is valid.
We are also interested in discovering a subset of words 
that relate persons to each other such as \RL{bin} or ``the son of''
or words that mean narrate such as \RL{qaal} or ``said''. 
With Sarf, we successfully automated the analysis of 
three books of Islamic narrations~\cite{IbnHanbal,AlTousi,AlKulayni}
and extracted from each one of them a vector of complex structures 
that are three level deep in hierarchy.
%We also showcase that the results of the accurate high level 
%abstraction
%results of the case-based analysis was not affected by a major
%refinement to the analyzer that increased low level analysis accuracy.

@@

Sarf is similar to Beesley~\shortcite{Beesley:01} in that it uses
finite-state Automata (FSA). 
Beesley reports that the number of the Automata generated by a compiler
from Xerox rules can not be controlled and also reports a difficulty to 
compose the FSAs into a single FSA framework. 
Sarf solves the problem with an FSA that is constructed from 
the controller and 
the prefix, stem, and suffix deterministic transducers. 
Sarf differs in that it uses recursive affixes in the 
affix transducers, and in that the transducers consider
input from the controller. 
With Sarf it is easy to control the FSA with a custom user case 
controller. Up to our understanding, Beesley  
requires recompiling the Xerox rules.

Sarf is similar to The MADA+TOKAN toolkit in that it thinks of
the several valid morphological analyses as a richness that 
should be exploited at a higher abstract level rather than
an inaccuracy that should be corrected. 
Sarf differs in that it provides an interface for the 
case study to interfere in judging the analyses much earlier and
at a finer granularity level. 
In Sarf, there is no need for a separate tokenizer such as
TOKAN since each solution keeps a stack of positions
that partition text into morphemes.
%The case-based controller can build the trace 
%when needed.

@@

Our approach is close to the refinement of SAMA~\cite{Maamouri:10}.
SAMA was refined to interact with
the ATB~\cite{Maamouri:04} project after the addition of a large 
new corpus. 
%We find supporting evidence in this decision to our hypothesis
%about the utility of a case-based analyzer. 
The algorithmic changes in SAMA were
done manually and worked in integration with the ATB format. 
Sarf is a case-based controller approach that can modify 
the behavior of the analyzer on the fly and can interact
with any case study. 
In Sarf, the analyzer remains general, and the 
case-based controller specializes the analysis.

@@

Sarf encodes the prefix, suffix and stem lexicons into
three linear transducers and uses a non-deterministic construction
of the three to compute all valid morphological analyses.
We first present a running example of Sarf.
The running example explains how Sarf works,
implements recursive affixes, and
handles ``run-on'' words.
%Then we discuss how Sarf builds and composes the different FSAs.

@@

\transfalse
\begin{figure}[tb]
\center{
\resizebox{\columnwidth}{!}
{ \input{figs/FSM.pdftex_t}}
\caption{Example affix and stem transducers.}
\label{f:example}
}
\end{figure}
% the FSMs for the three words
\transtrue

%{\bf Sarf example.}

The diagram in Figure~\ref{f:example}
illustrates the operation of $\Phi$ the Sarf FSA 
when parsing the
string \RL{al-lA`bwn}\RL{wsyl`b-h-aa}.
\footnote{ 
    \RL{n}%
        \RL{w}
    \RL{b}
    \RL{`}
    \RL{a} 
    \RL{l}
    \RL{l}
    \RL{a} %
        \RL{a}
    \RL{h}
    \RL{b}
    \RL{`}
    \RL{l}
    \RL{y}
    \RL{s}
    \RL{w}
    in separate form to ease following the example
}

The box and circle denote accept and regular
states respectively.
The edges are transitions with labels corresponding to
the input letters.
The reject states are omitted for clarity and are denoted
by the absence of corresponding transitions.

Subfigures (a), (b), and (c) in Figure~\ref{f:example}
represent ${\cal P}$ the prefix ,
${\cal S}$ the stem, and ${\cal X}$ the suffix
transducers respectively. 
The symbol $\epsilon$ represents an empty string and is 
the source of non-determinism in this example.

We start with $\Phi$ at the root square state in 
${\cal P}$ which is an accept state. 
The $\epsilon$-edge connects the ${\cal P}$
to ${\cal S}$ and transitions from any accept state
in ${\cal P}$ to the root state in ${\cal S}$.
The $\epsilon$-edge that connects the ${\cal S}$ to ${\cal X}$
follows the same behavior.

When there are two valid transitions such as \RL{w} 
and $\epsilon$ in the start case, 
$\Phi$ spawns an exact copy of itself $\Psi$. 
$\Phi$ makes one transition, and $\Psi$ makes the other. 
Each of $\Phi$ and $\Psi$ now represent a valid analysis so far. 
An FSA ($\Phi$ or $\Psi$) dies when it reaches a reject state.
In our example, if there were no stems that start 
with the letter \RL{w}, $\Psi$ will die. 
In reality, $\Psi$ will die when the input 
is at \RL{wsyl`}.

Note that the affix transducers allow recursive 
affixes. 
For example, \RL{w} will result in an accept state
that transitions to the ${\cal S}$.
When \RL{s} follows, we move to another accept state in 
${\cal P}$ corresponding to the prefix \RL{ws}. 
The same applies to ${\cal X}$.

Lets consider $\Phi$ after it consumed \RL{wsy} 
and transitioned into the root node of ${\cal S}$.
Now $\Phi$ will transition with \RL{l`b} to reach an accept 
state.

Before moving with the letter \RL{b} to the accept state,
$\Phi$ needs to make sure that the stem \RL{l`b} is compatible
with the prefix \RL{wsy}. 
It does that by associating a category value for each accept state. 
Sarf considers the value of the category as part of the state.
Thus each state shown in Figure~\ref{f:example} actually represents
more than one state. 
If the category of \RL{l`b} is compatible with the category of
\RL{wsy} then \RL{b} will move $\Phi$ to an accept state. 
Otherwise, it will move it to a regular or a reject state.

Since $\Phi$ is now in an accept state in ${\cal S}$, it will
spawn $\Xi$ that transitions into the root of ${\cal X}$. 
When \RL{h} is consumed, $\Phi$ will die since 
there is no edge from the current state labeled with \RL{h}. 
Now $\Xi$ represents a valid analysis and can proceed.

Note that $\Xi$ reports a full word at any accept state
by spawning a new FSA using the $\epsilon$ transition
to ${\cal P}$.
In Sarf, this only happens with white space, delimiters, 
and non-connecting letters. 
We left that out in Figure~\ref{f:example} for clarity.

@@

\begin{figure}[tb]
\center{
\resizebox{\columnwidth}{!}
{ \input{figs/abstract_machine.pdftex_t}}
\caption{Sarf NDFSA.}
\label{f:composition}
}
\end{figure}

@@

\begin{table}[tb]
%\centering
%\resizebox{.9\columnwidth}{!}{
\begin{tabular} {p{6cm}}
\begin{Verbatim}[fontsize=\relsize{-2},
frame=topline,framesep=4mm,label=\fbox{NDSarf algorithm},
commandchars=\\\{\}, codes={\catcode`$=3\catcode`_=8}]
NDSarf(string $L$, Controller $A$) 
  MachineVec $M$; -- collection of machines
  Machine $\Phi_1$;
  $M$.insert($\Phi_1$);
  foreach $c$ in $L$ \{
    foreach $\Phi$ in $M$ \{
      if ( $\Phi$.state.isAccept() ) \{
        if ($\Phi$.isSuffix())
          if ($c$.isWhite()) or 
            $A$.report();
          if ($\Phi$.lastChar().isNonConnecting())
            $A$.report();
        $\Psi$ = $\Phi$.clone();
        $M$.insert($\Psi$); \}

      if ( $\Phi$.isWalkable($c$) ) \{
        $\Phi$.transition($c$);
      \} else \{
        $M$.remove($\Phi$);
        $\Phi$.die();
      \} \} 
    $A$.control($M$, $c$); \}
\end{Verbatim}
\end{tabular}
%}
\label{a:ndsarf}
\end{table}

@@

Intuitively, the controller reports a valid chain 
of narrators when a sequence of names
connected by narrator connectors appears. 
It marks the beginning of the hadith with the beginning of 
the current sequence,
and marks the end of the hadith with the beginning of the 
next sequence. 
It also marks the chain of narrators as the sequence itself.

@@

@@

\begin{table*}[bt]
\centering
\caption{Results of the hadith case study with Sarf.}
%\begin{tabular}{|p{1.5cm}||c|c||c|c||c|c|} \hline
\resizebox{1.8\columnwidth}{!}{
\begin{tabular}{lp{.2cm}ccp{.2cm}ccp{.2cm}cc} %\cline{2-10}
 & & \multicolumn{2}{c}{Al Kafi} & & \multicolumn{2}{c}{Al Istibsar} & &\multicolumn{2}{c}{Ibn Hanbal} \\ \cline{3-10}
 & & Simple & Refined & & Simple & Refined & & Simple  & Refined \\\bottomrule  %\hline \hline
Word count & &\multicolumn{2}{c}{98,943} & & \multicolumn{2}{c}{103,835} & & \multicolumn{2}{c}{20,354} \\ 
 Names      & & 20,473 & 12,060  & &20,646 &  14,613& & 4,762 & 3,013\\
Names/Narrator& &2.48 & 1.97 & & 2.17 &  1.84& & 1.49 & 1.25 \\
Narrators & &3,080 & 2,623 & & 6,164 &  5,767& & 2,082 & 1,755 \\ 
Narrators/Chain & & 5.28 & 4.84 & & 4.98&  4.76 & & 4.8 &4.05 \\
Chains & & 583 & 542 &  & 1,238& 1,211& & 433 & 433 \\ 
Ignored names & & 11,801 & 6,400 &  & 5,845 & 3,348 & & 1,386 & 642 \\ \hline
Segmentation accuracy & & 85\% & 96\%& & 96\%& 96\%& & 92\%& 92\%\\ 
Chain accuracy & & 99\% & 99\%&  & 98\%& 99\%& & 99\% & 97\% \\ 
%Narrator accuracy & & & & & &  & & & \\ 
Narrator name accuracy & & 93\% & 91\%& &  92\%& 90\% & & 90\% & 90\% \\ \hline
Name false positives & & 28\% & 7\%&  & 9\%& 4\% & & 34\% & 4\% \\ \hline
Sarf running time (secs.)& & 2.97 & 1.32 & & 2.93 & 1.31 & & .25 & .096\\ \hline \hline
 & & Buckwalter& ElixirFM & & Buckwalter & ElixirFM & & Buckwalter & ElixirFM \\\bottomrule
Names      & & 18,354 & 19,201 & & 17,934 &  18,202 & & 5,021 & 5,117 \\
Names/Narrator& & 2.6 & 2.1 & & 2.47 &  2.07 & & 1.83 & 1.62\\
Narrators & & 3,210 & 2,981 & & 6,425 &  6,208 & & 2,311 & 1,965 \\ 
Narrators/Chain & & 5.57 & 5.32 & & 5.46 &  5.32 & & 5.12 & 5.01 \\
Chains & & 627 & 601 &  & 1,318 & 1,250 & & 487 & 452 \\ 
Ignored names & & 9,642 & 9,915 &  & 6,118 & 6,874 & & 1,812 & 1,904 \\ \hline
Segmentation accuracy & & 76\%& 84\%& & 85\%& 89 \%& & 84\%& 85\%\\ 
Chain accuracy & & 78\% & 87\%&  & 87\%& 90\%& & 85\% & 86\% \\ 
%Narrator accuracy & & & & & &  & & & \\ 
Narrator name accuracy & & 76\% & 86\%& & 84\%& 87\% & & 84\% & 84\% \\ \hline
Name false positives & & 29\% & 22\%&  & 14\%& 13\% & & 36\% & 32\% \\ \hline
Running time (secs.)& & 6.65 & 2.78$\times60^2$& &4.55 & 2.3$\times60^2$ & & .66 & 29.2$\times$60 
\end{tabular}
}
\normalsize
\label{t:hadithresallresults}
\end{table*}

In this section we present results that compare Sarf to existing 
morphological analyzers such as Buckwalter and ElixirFM. 
We also evaluate our hypothesis of the utility of a case-based 
morphological analyzer.
We validate that the high-level analysis of the output 
of low-level morphological analysis can compensate for accuracy.
To achieve that we ran Sarf in two modes, simple and refined.
In the simple mode, we used Sarf as discussed in 
Section~\ref{sec:sarf} with the 
controller of Section~\ref{sec:controller}.
In the refined mode, we applied several low-level morphological 
refinements to Sarf such as disambiguation of 
female proper names so that the controller can ignore them. 
We also added a refinement that detected compound proper names 
such as \RL{`bd alr.hmn} 
as they are common in narrator names.
%Finally, we consider a possessive form as a reference to 
%a person and treat
%it as a narrator.

We ran our experiments using a dual core 2.66 Ghz 64-bit processor 
with 4GB of memory using a Linux operating system.

We report our results in Table~\ref{t:hadithresallresults} on the 
three hadith books.
In general, we observe that the refinements slightly 
improved performance and accuracy at
the lowest level of abstraction by reducing false positives.
However, the refinements had little to no effect on the high 
accuracy measures at higher levels of abstraction.
We believe that this is a strong supporting evidence to our 
hypothesis.

The row for ``word count'' is an indicator of the size of the data.
The %numbers in 
rows ``names'', ``narrators'', and ``chains'' denote the total
number of the corresponding entity per book.
The relative magnitudes of these rows
reflect the hierarchical structure where 
a chain contains
narrators, and a narrator contains names.
The row ``ignored names'' reports the number of names detected 
outside the chains and thus are irrelevant to
the case study.
In addition, the rows ``names/narrator'' and ``narrators/chain'' 
report the average
number of names per narrator and 
the average number of narrators per chain respectively.

The segmentation accuracy reports on the accuracy of 
segmentation of the book into separate hadiths.
The metric is computed manually over a representative sample 
of each book.
The same applies to all other accuracy measures.
The chain accuracy reflects the percentage of correct narrators 
covered by extracted chains.
The ``narrator name accuracy'' row reflects the percentage of 
correct proper names detected within
narrator structures.

The row ``name false positives'' reports the percentage of words 
recognized as names while they should not.
We notice that this low level metric improved dramatically 
with the refined analysis without substantially affecting the 
high abstract measures such as chain and segmentation
accuracy.

We notice that the controller in the simple analysis ignored 
more names and thus the case-based
controller compensated with a simple heuristic
for the inaccuracy of the low level 
analysis.
%and without the need of the low-level refinements.

We investigated the case of Al-Kafi where the accuracy of 
the simple analysis was significantly lower than that of the 
refined analysis. 
We discovered that the chains detected by the refined analysis 
were a subset of those detected by the simple
analysis. 
The reason for the lower accuracy is false positive chains, 
implying that the simple analysis did not miss important 
information.

We hypothesize that a higher level of abstraction such as 
intersecting the narrators from the hadith books with the narrators 
in the biographies can reduce the false positives in the chains
in this case.

@@

We implemented our case study with Buckwalter in two phases. 
In phase one, we processed the entire text with Buckwalter and
obtained annotations to text. 
In the second phase, we fed the annotations to 
the hadith controller as if they were coming from Sarf. 
We measured the time it took to produce the annotations
and the time it took the controller to process the 
annotations without the time spent streaming the annotations.
We used the same setup for ElixirFM.

Sarf outperforms both Buckwalter and ElixirFM in terms of 
running time. 
It also produces higher accuracy measures. 
We think this is mainly due to the extended
lexicon as well as to the fact that Sarf reduces false positive
names by ignoring prefixes and suffixes that can not form
names even if one possible alternative of the stem is a name. 
Doing the same with Buckwalter and ElixirFM requires  
expanding the annotations resulting from one simple affix to be 
reflected in all the composite affixes that contain it.
%expanding the annotations resulting from one affix to reflect
%its multiple composing affixes.
%However, this was easily performed using our recursive definition 
%of the affixes.
